---
title: "Savannah, GA Analysis"
output: html_notebook
---
This notebook will deep dive into the Savannah, GA metro area. The goal of the analysis is to provide an overview of the metro area and discuss certain areas that are worth exploring further.

The analysis will be split up into a few sections:

1. Metro Area Summary
2. Submarket Screening
3. Spillover and Adjacency Analysis
4. Risk and Demand Analysis
5. Scoring and Prioritization of sub-Geos
6. Summary

# Load the enviroment

In this section we will load required libraries, ingest datasets, set functions.

```{r Load Libraries}
library(tidyverse)
library(lubridate)
library(magrittr)
library(plotly)
library(tidytext)
library(scales)
library(tidycensus)
library(tigris)
library(sf)
library(readxl)
library(httr)
library(jsonlite)
library(DT)
library(ggbeeswarm)
library(knitr)
library(factoextra)
library(corrplot)
library(leaflet)
library(ggrepel)
```

```{r Load Custom Functions}
# Data Cleaning
source("/Users/danberle/Desktop/real_estate_investment/functions/functions_data.R")
# Data Viz
source("/Users/danberle/Desktop/real_estate_investment/functions/functions_viz.R")
```

```{r Ingest Base Tables}

```

# Metro Area Summary

This section will review the overall performance of the metro area. This section will have 3 main areas:

1. Overall Summary Stats
  - Region
  - Population
  - Indexes
  - Education
2. Comparisons
  - Index ranking
  - Population ranking
  - Economic Growth ranking
  - Housing Growth ranking
3. KPI Growth
  - Population Growth
  - Economic Growth
  - Housing Growth

This section presents an overall summary of the Savannah, GA Metro Area and compares the Metro Area to National and Regional Averages


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

