---
title: "Southeastern US Analysis"
output: html_notebook
---

This notebook analyzes key demographic figures of metro areas in the Southeastern US. Our goal is to identify the fastest growing metro areas by population growth and explore a few related variables

We will follow this structure to our analysis:
  1. Ingest our data sources
  2. Calculate core KPIs
  3. Generate rankings & visuals
  4. Discuss main takeaways

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*. 

# Set up environment
```{r Load Packages, include - FALSE}
library(tidyverse)
library(lubridate)
library(magrittr)
library(plotly)
library(tidytext)
library(scales)
library(tidycensus)
library(tigris)
library(sf)
library(readxl)
library(httr)
library(jsonlite)
library(DT)
library(ggbeeswarm)
library(knitr)
library(factoextra)
library(corrplot)
library(leaflet)
library(ggrepel)
library(slider)
library(gt)
library(ragg)
```

```{r Load custom functions, include = FALSE}
# Data Cleaning
source("/Users/danberle/Documents/projects/real_estate_investment/functions/functions_data.R")
# Data Viz
source("/Users/danberle/Documents/projects/real_estate_investment/functions/functions_viz.R")
```

```{r Create new functions, include = FALSE}
# Generic helper to add simple % change and CAGR for multiple horizons
# df must contain: id_cols (e.g., GEOID, NAME), a YEAR column (numeric/int), and a value_col
add_growth_cols <- function(df, id_cols, year_col = "YEAR", value_col, horizons = c(1,3,5,10), prefix = "") {
  stopifnot(all(c(id_cols, year_col, value_col) %in% names(df)))
  df %>%
    arrange(across(all_of(c(id_cols, year_col)))) %>%
    group_by(across(all_of(id_cols))) %>%
    mutate(
      across(
        all_of(value_col),
        .fns = list(
          !!!setNames(
            lapply(horizons, function(h) \(x) (x / dplyr::lag(x, h) - 1)),
            paste0(prefix, "chg_", horizons, "y")
          ),
          !!!setNames(
            lapply(horizons, function(h) \(x) (x / dplyr::lag(x, h))^(1 / h) - 1),
            paste0(prefix, "cagr_", horizons, "y")
          )
        ),
        .names = "{fn}"
      )
    ) %>%
    ungroup()
}
```

```{r Ingest GEO Data, include=FALSE}
cbsa_shapes <- core_based_statistical_areas(cb = TRUE, year = 2023) %>%
  st_transform(crs = 5070)

# Load state and county outlines
states_sf <- states(cb = TRUE, year = 2023) %>% st_transform(4326)
counties_sf <- counties(cb = TRUE, year = 2023) %>% st_transform(4326)

# Optional: Filter to lower 48
states_sf <- states_sf %>% filter(!STATEFP %in% c("02", "15", "72", "78", "66", "69", "60"))
counties_sf <- counties_sf %>% filter(!STATEFP %in% c("02", "15", "72", "78", "66", "69", "60"))
```

# Load Data

The main data we are using for this analysis and their sources are:
  1. Population Growth - ACS (PEP Future)
  2. Median Income, Median Age, Housing Units - ACS
  3. Housing Permits Per Capita (Last 5 Years) - BPS
  4. GDP Growth - BEA
  5. Unemployment Rate - BLS
  
```{r Ingest}
# Metro Metadata
metro_metadata <- read_csv("/Users/danberle/Documents/projects/real_estate_investment/data/gold/cbsa/cbsa_metadata.csv") %>%
  mutate(GEOID = as.character(GEOID))

# ACS
acs_silver <- read_csv("/Users/danberle/Documents/projects/real_estate_investment/data/silver/cbsa/acs_cbsa_5_year.csv")

# BPS
bps_silver <- read_csv("/Users/danberle/Documents/projects/real_estate_investment/data/silver/cbsa/bps_cbsa.csv")

# GDP
bea_silver <- read_csv("/Users/danberle/Documents/projects/real_estate_investment/data/silver/cbsa/bea_cbsa_gdp_summary.csv")

# BLS
bls_silver <- read_csv("/Users/danberle/Documents/projects/real_estate_investment/data/silver/cbsa/bls_cbsa_unemployment.csv")
```

# Calculate KPIs

In this section we will clean our data sets and calculate our KPIs

We start by calculating population growth by metro area. We calculate 1, 3, 5, and 10 year population growth rates. 
```{r Population Growth Metrics}
# Pop Growth
metrics_acs_pop_growth <- acs_silver %>%
  select(GEOID, NAME, YEAR = year, TOTAL_POP = pop_totalE) %>%
  mutate(YEAR = as.integer(YEAR)) %>%
  add_growth_cols(
    id_cols = c("GEOID", "NAME"),
    year_col = "YEAR",
    value_col = "TOTAL_POP",
    horizons = c(1,3,5,10),
    prefix = "pop_") %>%
  mutate(GEOID = as.character(GEOID))

# Income Growth
metrics_acs_income_growth <- acs_silver %>%
  select(GEOID, NAME, YEAR = year, MEDIAN_INCOME = income_medE) %>%
  mutate(YEAR = as.integer(YEAR)) %>%
  add_growth_cols(
    id_cols = c("GEOID", "NAME"),
    year_col = "YEAR",
    value_col = "MEDIAN_INCOME",
    horizons = c(1,3,5,10),
    prefix = "income_") %>%
  mutate(GEOID = as.character(GEOID))
```

Next, we select our main demographic KPIs from the ACS: Income, Age, Housing Units
```{r Demographic Metrics}
metrics_acs_demographics <- acs_silver %>%
  select(GEOID, NAME, YEAR = year, MEDIAN_INCOME = income_medE, MEDIAN_AGE = age_medE, HOUSING_UNITS = total_housing_unitsE, VACANCY_RATE = vacancy_rate, HIGHER_ED_RATE = ed_higher_pop_percent, LFP_RATE = labor_force_participation_rate) %>%
  mutate(GEOID = as.character(GEOID))
```

We follow this up by calculating our housing permit metrics from the BPS. We will start by rolling up our monthly metrics to a yearly level. We then calculate a per capita rate by taking units per 1000 people, doing this for the most recent year and taking a 5 year average of units per current population. Lastly, we calculate our unit growth rate. 

```{r Building Units}
# Annualize BPS
  # We need to add leading zeroes to GEOID
bps_annual <- bps_silver %>%
  select(GEOID = CBSA_CODE, YEAR, MONTH, TOTAL_BLDGS, TOTAL_UNITS) %>%
  mutate(YEAR = as.integer(YEAR)) %>%
  group_by(GEOID, YEAR) %>%
  summarise(
    permits_units = sum(TOTAL_UNITS, na.rm = TRUE),
    permits_buildings = sum(TOTAL_BLDGS, na.rm = TRUE),
    .groups = "drop"
  )

# Join population to normalize (use the same YEAR in pop_kpis)
bps_with_pop <- bps_annual %>%
  left_join(
    metrics_acs_pop_growth %>% 
      select(GEOID, YEAR, TOTAL_POP),
    by = c("GEOID", "YEAR")
  )

# Compute normalized KPIs + 5-year rolling averages (centered on YEAR)
metrics_bps_kpis <- bps_with_pop %>%
  group_by(GEOID) %>%
  arrange(GEOID, YEAR) %>%
  mutate(
    units_per_1k = if_else(TOTAL_POP > 0, 1000 * permits_units / TOTAL_POP, NA_real_),
    bldgs_per_1k = if_else(TOTAL_POP > 0, 1000 * permits_buildings / TOTAL_POP, NA_real_),
    units_per_1k_5y_avg = slider::slide_dbl(units_per_1k, mean, .before = 4, .complete = TRUE, na.rm = TRUE),
    bldgs_per_1k_5y_avg = slider::slide_dbl(bldgs_per_1k, mean, .before = 4, .complete = TRUE, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  # Also add growth rates on annual units if helpful
  add_growth_cols(
    id_cols = c("GEOID"),
    year_col = "YEAR",
    value_col = "permits_units",
    horizons = c(1,3,5),
    prefix = "permits_units_"
  )

# Typical fields you’ll use downstream:
# - units_per_1k_5y_avg (primary)
# - units_per_1k (latest year)
# - permits_units_chg_5y / permits_units_cagr_5y (optional)
```

We follow up our BPS data with calculations of our GPD growth per CBSA. Here, we will go for CAGR and Simple Growth Rates using GDP Chained 2017.
```{r GDP}
metrics_gdp_kpis <- bea_silver %>%
  select(GEOID = GeoFIPS, YEAR = Year, GDPChained2017, GDPCurrentDollars) %>%
  mutate(YEAR = as.integer(YEAR)) %>%
  add_growth_cols(
    id_cols = c("GEOID"),
    year_col = "YEAR",
    value_col = "GDPChained2017",
    horizons = c(1,3,5,10),
    prefix = "gdp_real_"
  ) %>%
  # Optional: also compute for current-dollar
  add_growth_cols(
    id_cols = c("GEOID"),
    year_col = "YEAR",
    value_col = "GDPCurrentDollars",
    horizons = c(1,3,5,10),
    prefix = "gdp_nom_"
  )
```

To finish up our KPI calcualtions we calculate unemployment rates per Metro using our LAUS data from BLS.
```{r Unemployment}
metrics_laus_kpis <- bls_silver %>%
  mutate(YEAR = as.integer(year),
         GEOID = as.character(substr(area_code, 5, 9))
         ) %>%
  select(GEOID, YEAR, unemployment_rate) %>%
  group_by(GEOID) %>%
  arrange(GEOID, YEAR) %>%
  mutate(
    unemp_rate_chg_5y = unemployment_rate - dplyr::lag(unemployment_rate, 5),
    unemp_rate_chg_10y = unemployment_rate - dplyr::lag(unemployment_rate, 10)
  ) %>%
  ungroup()
```

To wrap up the section, we will join together our metrics tables, select the appropriate year and select a final list of KPIs
```{r Final Metrics Table}

# Select the latest year we will use for the analysis
latest_year <- 2023L

# Create the most recent metrics tables
metrics_latest_pop <- metrics_acs_pop_growth %>%
  filter(YEAR == latest_year) %>%
  select(GEOID, YEAR,
         TOTAL_POP,
         starts_with("pop_chg_"), starts_with("pop_cagr_"))

metrics_latest_income <- metrics_acs_income_growth %>%
  filter(YEAR == latest_year) %>%
  select(GEOID, YEAR,
         MEDIAN_INCOME,
         starts_with("income_chg_"), starts_with("income_cagr_"))

metrics_latest_demo <- metrics_acs_demographics %>%
  filter(YEAR == latest_year) %>%
  select(GEOID, YEAR,
         MEDIAN_AGE, HOUSING_UNITS, VACANCY_RATE, HIGHER_ED_RATE, LFP_RATE)

metrics_latest_bps <- metrics_bps_kpis %>%
  filter(YEAR == latest_year) %>%
  select(GEOID, YEAR,
         units_per_1k, units_per_1k_5y_avg,
         starts_with("permits_units_")) # optional growth fields

metrics_latest_gdp <- metrics_gdp_kpis %>%
  filter(YEAR == latest_year) %>%
  select(GEOID, YEAR,
         GDPChained2017, GDPCurrentDollars,
         starts_with("gdp_real_"), starts_with("gdp_nom_"))

metrics_latest_laus <- metrics_laus_kpis %>%
  filter(YEAR == latest_year) %>%
  select(GEOID, YEAR, unemployment_rate, unemp_rate_chg_5y, unemp_rate_chg_10y)

# Create our final join using our CBSA Metadata as a base
final_cbsa_kpis <- metro_metadata %>%
  select(GEOID, NAME, primary_state, region, division, states, cbsa_type) %>%
  left_join(metrics_latest_pop, by = "GEOID") %>%
  left_join(metrics_latest_income %>% select(-YEAR), by = "GEOID") %>%
  left_join(metrics_latest_demo %>% select(-YEAR), by = "GEOID") %>%
  left_join(metrics_latest_bps %>% select(-YEAR), by = "GEOID") %>%
  left_join(metrics_latest_gdp %>% select(-YEAR), by = "GEOID") %>%
  left_join(metrics_latest_laus %>% select(-YEAR), by = "GEOID") %>%
  filter(!is.na(YEAR))
```

# Insights & Rankings

Now that we have our core KPIs we will calculate rankings of the metro areas and begin to explore the data. The most important rankings to start with is 5 year population as this takes into account a two years before COVID and 3 years post.

We will also focus on 5 Year GDP Growth, Current Unemployment, Units Per 1k (5 Year Avg), Vacancy Rate, Median Income, and Median Age

```{r Create rankings and select variables}
# Set our ranking metric
rank_metric <- "pop_cagr_5y"

# Select a subset of KPIs
analysis_kpis <- final_cbsa_kpis %>%
  select(GEOID:TOTAL_POP, pop_cagr_5y, MEDIAN_INCOME:LFP_RATE, units_per_1k_5y_avg, gdp_real_cagr_5y, unemployment_rate) 

analysis_cbsa_growth_ranks <- analysis_kpis %>%
  filter(YEAR == latest_year, !is.na(.data[[rank_metric]]), !is.na(cbsa_type)) %>%
  # overall national rank across all CBSAs (Metro+Micro)
  mutate(
    nat_rank_all = min_rank(desc(.data[[rank_metric]])),
    nat_pct_all  = percent_rank(.data[[rank_metric]])
  ) %>%
  # national rank *within class*
  group_by(cbsa_type) %>%
  mutate(
    nat_rank_in_cbsa_type = min_rank(desc(.data[[rank_metric]])),
    nat_pct_in_cbsa_type  = percent_rank(.data[[rank_metric]])
  ) %>%
  ungroup() %>%
  # regional flag
  mutate(in_region = region == "South") %>%
  # regional rank *within class*
  group_by(cbsa_type, in_region) %>%
  mutate(
    reg_rank_in_cbsa_type = min_rank(desc(.data[[rank_metric]])),
    reg_pct_in_cbsa_type  = percent_rank(.data[[rank_metric]])
  ) %>%
  ungroup()

analysis_south_growth_ranks <- analysis_cbsa_growth_ranks %>%
  filter(region == "South") %>%
  arrange(reg_rank_in_cbsa_type) %>%
  mutate(
    # 5 quantile bins: 5 = fastest growth
    growth_q5 = ntile(.data[[rank_metric]], 5),
    growth_q5_lbl = factor(growth_q5,
      levels = 1:5,
      labels = c("Bottom 20%", "20–40%", "40–60%", "60–80%", "Top 20%"))
  )

```

```{r Join Rankings DF to CBSA Shapefile}
analysis_shp_cbsa_growth_ranks <- cbsa_shapes %>%
  left_join(analysis_south_growth_ranks %>% select(!NAME), by = c("GEOID" = "GEOID")) %>%
  filter(region == "South")
```

Next, we will start answering some core questions

# *Overview*
In this section we will review metro area growth rates and highlight the top 10. 

Regional Map of KPIs, colored by the growth ranking (grwoth label will be our cluster) and a bar chart of our top 10 markets, with a nicely formatted table of key metrics

```{r CBSAs Top Growth Map}
# --- Inputs you already have ---
metric_var    <- "pop_cagr_5y"
rank_var      <- "reg_rank_in_cbsa_type"
top_n_labels  <- 10
alpha_fill    <- 0.9
se_sc_states  <- c("FL","GA","SC","NC","TN","AL","MS","AR","LA","TX","OK","VA","WV","KY")

# --- 1) Filter states, set a common CRS ---
states_se <- states_sf %>% filter(STUSPS %in% se_sc_states)
plot_crs  <- st_crs(states_se)

# transform *and* filter CBSA layer to match states before plotting
cbsa_sf <- analysis_shp_cbsa_growth_ranks %>%
  filter(primary_state %in% se_sc_states, 
         !is.na(.data[[metric_var]]),
         cbsa_type == "Metro Area") %>%
  st_transform(plot_crs)

states_se <- st_transform(states_se, plot_crs)

# --- 2) Top-N labels computed in the same CRS as the map ---
labs_df <- cbsa_sf %>%
  filter(!is.na(.data[[rank_var]])) %>%
  slice_min(order_by = .data[[rank_var]], n = top_n_labels, with_ties = TRUE) %>%
  mutate(
    label_pt = st_point_on_surface(geometry),
    coords   = st_coordinates(label_pt),
    x = coords[,1],
    y = coords[,2]
  )

topN_outline <- labs_df  # same Top-N for bold outline

# --- 3) Optional: set extent from states bbox ---
bb   <- st_bbox(states_se)
xlim <- c(bb["xmin"], bb["xmax"])
ylim <- c(bb["ymin"], bb["ymax"])

# --- 4) Plot ---
p <- ggplot() +
  geom_sf(data = states_se, fill = NA, color = "grey40", linewidth = 0.4) +

  # main CBSA fill
  geom_sf(data = cbsa_sf, aes(fill = .data[[metric_var]]),
          color = "white", linewidth = 0.25, alpha = alpha_fill) +
  # Add double stroke for state highlighting
  geom_sf(data = states_se, fill = NA, color = "white", linewidth = 0.7) +
  geom_sf(data = states_se, fill = NA, color = "grey30", linewidth = 0.35) +
  # subtle dark edge on all CBSAs
  geom_sf(data = cbsa_sf, fill = NA, color = "grey25", linewidth = 0.18) +
  # bold edge for Top-N CBSAs
  geom_sf(data = topN_outline, fill = NA, color = "black", linewidth = 0.45) +

  # labels (computed in the same CRS, using numeric x/y)
  ggrepel::geom_label_repel(
    data = labs_df, aes(x = x, y = y, label = NAME),
    size = 3, seed = 123,
    fill = scales::alpha("white", 0.75),
    label.size = 0.2,
    min.segment.length = 0,
    max.overlaps = Inf
  ) +

  scale_fill_viridis_c(
    option = "viridis",
    name   = "Population CAGR (5y)",
    labels = ~ scales::percent(.x, accuracy = 0.1),
    guide  = guide_colorbar(barheight = unit(3.5, "cm"))
  ) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE, clip = "on") +
  labs(
    title = "Southeast/South Central — CBSA Population Growth",
    subtitle = "Fill: Pop CAGR (5y). Labels & bold outline: Top 10 metros"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.title = element_text(face = "bold")
  )

p
```

```{r Quantiles Map}
metric_var    <- "pop_cagr_5y"
rank_var      <- "reg_rank_in_cbsa_type"
top_n_labels  <- 10
alpha_fill    <- 0.9
se_sc_states  <- c("FL","GA","SC","NC","TN","AL","MS","AR","LA","TX","OK","VA","WV","KY")

# --- data prep (same as before; ensure common CRS & SE subset) ---
states_se <- states_sf %>% filter(STUSPS %in% c("FL","GA","SC","NC","TN","AL","MS","AR","LA","TX","OK","VA","WV","KY"))
plot_crs  <- st_crs(states_se)

cbsa_sf <- analysis_shp_cbsa_growth_ranks %>%
  filter(primary_state %in% states_se$STUSPS,
         !is.na(.data[[metric_var]]),
         cbsa_type == "Metro Area") %>%
  st_transform(plot_crs)

states_se <- st_transform(states_se, plot_crs)

# Compute quintile labels on the metric
cbsa_q <- cbsa_sf %>%
  mutate(
    growth_q5 = ntile(.data[[metric_var]], 5),
    growth_q5_lbl = factor(
      growth_q5,
      levels = 1:5,
      labels = c("Bottom 20%","20–40%","40–60%","60–80%","Top 20%")
    )
  )

# Top-N for labels/outline (same selection logic as continuous)
labs_df <- cbsa_q %>%
  filter(!is.na(.data[[rank_var]])) %>%
  slice_min(order_by = .data[[rank_var]], n = top_n_labels, with_ties = FALSE) %>%
  mutate(
    label_pt = st_point_on_surface(geometry),
    coords   = st_coordinates(label_pt),
    x = coords[,1], y = coords[,2]
  )

topN_outline <- labs_df

bb   <- st_bbox(states_se)
xlim <- c(bb["xmin"], bb["xmax"]); ylim <- c(bb["ymin"], bb["ymax"])

# --- plot (quintiles) ---
p_q <- ggplot() +

  # fill by quintile label
  geom_sf(
    data = cbsa_q,
    aes(fill = growth_q5_lbl),
    color = "white", linewidth = 0.22, alpha = alpha_fill
  ) +
  # Add double stroke for state highlighting
  geom_sf(data = states_se, fill = NA, color = "white", linewidth = 0.7) +
  geom_sf(data = states_se, fill = NA, color = "grey30", linewidth = 0.35) +
  # subtle dark edge
  geom_sf(data = cbsa_q, fill = NA, color = "grey25", linewidth = 0.16) +
  # bold outline for Top-N
  geom_sf(data = topN_outline, fill = NA, color = "black", linewidth = 0.45) +

  # labels
  ggrepel::geom_label_repel(
    data = labs_df, aes(x = x, y = y, label = NAME),
    size = 3, seed = 123,
    fill = alpha("white", 0.75), label.size = 0.2,
    min.segment.length = 0, max.overlaps = Inf
  ) +

  # discrete viridis (nice for ordered categories)
  scale_fill_viridis_d(
    option = "viridis",
    direction = 1,
    name = "Pop Growth Quintile"
  ) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE, clip = "on") +
  labs(
    title = "Southeast/South Central — CBSA Population Growth (Quintiles)",
    subtitle = "Fill: Quintiles of Pop CAGR (5y). Bold outline & labels: Top 10 metros"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(), axis.title = element_blank(),
    axis.ticks = element_blank(), panel.grid = element_blank(),
    legend.title = element_text(face = "bold")
  )

p_q
```

```{r Save the file}
# Best: set both plot and legend backgrounds to white
p_q_fixed <- p_q +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_growth_quintiles.png", p_q_fixed,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```

We follow this up Bar Charts showing our Top Markets and the growth rates by state

```{r Bar Chart Data Prep}
metric_var   <- "pop_cagr_5y"
rank_var     <- "reg_rank_in_cbsa_type"      # swap to "nat_rank" if needed
top_n_labels <- 20

# You can build from cbsa_sf (sf object) or from your non-sf final KPI table.
# Here I assume we start from cbsa_sf (already filtered to SE states).
top20_tbl <- cbsa_sf %>%
  st_drop_geometry() %>%
  filter(!is.na(.data[[metric_var]]), !is.na(.data[[rank_var]])) %>%
  slice_min(order_by = .data[[rank_var]], n = top_n_labels, with_ties = FALSE) %>%
  arrange(.data[[rank_var]]) %>%
  mutate(rank_display = row_number())
```

```{r Continious Bar Chart}
p_bar_cont <- top20_tbl %>%
  mutate(metro_lbl = factor(NAME, levels = rev(NAME))) %>%  # order by rank
  ggplot(aes(x = metro_lbl, y = .data[[metric_var]], fill = .data[[metric_var]])) +
  geom_col(width = 0.7, color = "grey25") +
  geom_text(aes(label = percent(.data[[metric_var]], accuracy = 0.1)),
            hjust = -0.05, size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_viridis_c(option = "viridis", name = "Pop CAGR (5y)",
                       limits = c(min(cbsa_sf$pop_cagr_5y, na.rm=TRUE),
                                  max(cbsa_sf$pop_cagr_5y, na.rm=TRUE))
             ) +
  coord_flip() +
  labs(
    title    = "Top 20 Fastest-Growing Metros (SE/South Central)",
    subtitle = "Ranked by 5-year Population CAGR",
    x = NULL, y = "Population CAGR (5y)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right",
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 9),
    plot.title = element_text(face = "bold")
  )

p_bar_cont
```
```{r Export Bar Chart}
# Best: set both plot and legend backgrounds to white
p_bar_cont_fixed <- p_bar_cont +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_growth_top20_markets.png", p_bar_cont_fixed,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```

We should update the color scale of our bar chart to be based on all values instead of the top 20 values we maintained


```{r Simple Table}
analysis_final_table <- analysis_shp_cbsa_growth_ranks %>%
  st_drop_geometry() %>%
  select(GEOID, NAME, primary_state, region, division, cbsa_type,
         TOTAL_POP, pop_cagr_5y, MEDIAN_INCOME, MEDIAN_AGE, VACANCY_RATE, 
         HIGHER_ED_RATE, LFP_RATE, income_cagr_5y,
         units_per_1k_5y_avg, gdp_real_cagr_5y, unemployment_rate,
         nat_rank_in_cbsa_type, reg_rank_in_cbsa_type, growth_q5_lbl) %>%
  mutate(unemployment_rate = unemployment_rate / 100)

write_csv(analysis_final_table, "/Users/danberle/Documents/projects/demographic_analysis/outputs/south_cbsa_final_kpis.csv")
```

```{r Publication Table}
top_n <- 20

tbl_top <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  slice_min(order_by = reg_rank_in_cbsa_type, n = top_n, with_ties = FALSE) %>%
  transmute(
    Rank   = reg_rank_in_cbsa_type,
    Metro  = NAME,
    State  = primary_state,
    `Pop CAGR (5y)`       = pop_cagr_5y,          # ensure these are proportions (e.g., 0.034, not 3.4)
    `Permits / 1k (5y)`   = units_per_1k_5y_avg,
    `Real GDP CAGR (5y)`  = gdp_real_cagr_5y,     # proportion
    `Unemp. Rate (2023)`  = unemployment_rate,    # proportion (e.g., 0.031)
    `Pop (2023)`          = TOTAL_POP
  ) %>%
  arrange(Rank)

gt_tbl <- tbl_top %>%
  gt() %>%
  fmt_percent(columns = c(`Pop CAGR (5y)`, `Real GDP CAGR (5y)`, `Unemp. Rate (2023)`),
              decimals = 1) %>%
  fmt_number(columns = c(`Permits / 1k (5y)`), decimals = 1) %>%
  fmt_number(columns = c(`Pop (2023)`), decimals = 0, use_seps = TRUE) %>%
  cols_align(align = "left", columns = c(Metro)) %>%
  cols_width(
    Rank ~ px(60),
    State ~ px(70),
    Metro ~ px(280),
    everything() ~ px(130)
  ) %>%
  tab_header(
    title = md("**Top 20 Southern Metros — Key Metrics**"),
    subtitle = "Population growth, housing pipeline, GDP, and labor market"
  ) %>%
  tab_options(
    table.font.size = px(12),
    data_row.padding = px(4),
    table.background.color = "white"
  )

gt_tbl
```

```{r Save the table}
gt::gtsave(gt_tbl, "/Users/danberle/Documents/projects/demographic_analysis/outputs/top20_south_metro_metrics.png",
           vwidth = 1400, vheight = 900, expand = 5, path = ".")
```


# *Supply & Demand*
Population Growth vs Permits - Scatter Plot

This section will explore the relationship between growth and permits. We will create a scatter plot to show the relationship between these variables.
Then, we will explore markets that seem to be under supplied, markets with high growth and high permitting. Lastly, we will look at any outliers.

```{r Housing vs Permits Scatter}
# ----- INPUTS (tweak these to your column names) -----
df_scatter <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  filter(!is.na(pop_cagr_5y), !is.na(units_per_1k_5y_avg))

metric_x   <- "pop_cagr_5y"        # proportion (e.g., 0.029 = 2.9%)
metric_y   <- "units_per_1k_5y_avg" # permits per 1,000 residents
size_var   <- "TOTAL_POP"           # population for point size
label_var  <- "NAME"
state_var  <- "primary_state"
rank_var   <- "reg_rank_in_cbsa_type" # regional rank within metros
top_n_lab  <- 20

# ----- MEDIANS: choose unweighted or weighted (by pop) -----
# Unweighted medians (simple, transparent)
x_med <- median(df_scatter[[metric_x]], na.rm = TRUE)
y_med <- median(df_scatter[[metric_y]], na.rm = TRUE)

# (Optional) Population-weighted medians:
# wtd_median <- function(x, w) x[order(x)][cumsum(w[order(x)])/sum(w) >= 0.5][1]
# x_med <- wtd_median(df[[metric_x]], df[[size_var]])
# y_med <- wtd_median(df[[metric_y]], df[[size_var]])

# ----- TOP 20 for labels & bold outline -----
top_lab <- df_scatter %>%
  arrange(!!sym(rank_var)) %>%
  slice_head(n = top_n_lab)

# Cap point sizes so large metros stand out but don't overwhelm
size_scaled <- function(x) {
  s <- scales::rescale(x, to = c(2.2, 10))  # adjust min/max point size here
  pmin(s, 10)
}

# Build four rectangles using the medians
xr <- range(df[[metric_x]], na.rm = TRUE)
yr <- range(df[[metric_y]], na.rm = TRUE)

# Build plot
p_scatter <- df_scatter %>%
  mutate(
    point_size = size_scaled(.data[[size_var]]),
    label_txt  = paste0(.data[[label_var]]),
    # Quadrant classification (optional; useful for tooltip/text)
    quadrant = case_when(
      .data[[metric_x]] >= x_med & .data[[metric_y]] >= y_med ~ "Balanced (High/High)",
      .data[[metric_x]] >= x_med & .data[[metric_y]] <  y_med ~ "Undersupplied (High/Low)",
      .data[[metric_x]] <  x_med & .data[[metric_y]] >= y_med ~ "Potentially Oversupplied (Low/High)",
      TRUE ~ "Stagnant (Low/Low)"
    )
  ) %>%
  ggplot(aes(x = .data[[metric_x]], y = .data[[metric_y]])) +

  # Base points
  geom_point(aes(size = point_size),
             color = "grey35", fill = "#2C7FB8", alpha = 0.7, shape = 21, stroke = 0.2) +

  # Bold outline for Top 20
  geom_point(data = top_lab,
             aes(x = .data[[metric_x]], y = .data[[metric_y]]),
             shape = 21, size = 3.6, stroke = 1.0, color = "black", fill = NA) +

  # Median quadrant lines
  geom_vline(xintercept = x_med, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_hline(yintercept = y_med, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  
    # Median value annotations
  annotate("label", x = x_med, y = -Inf,
           label = paste0("Median Pop CAGR: ", scales::percent(x_med, 0.1)),
           vjust = -0.4, size = 3, label.size = 0, fill = scales::alpha("white", 0.9)) +
  annotate("label", x = -Inf, y = y_med,
           label = paste0("Median Permits/1k: ", round(y_med, 1)),
           hjust = -0.05, size = 3, label.size = 0, fill = scales::alpha("white", 0.9)) +


  # Quadrant labels (subtle)
  annotate("label", x = -Inf, y =  Inf, label = "Low Growth / High Permits",
           hjust = -0.05, vjust = 1.05, size = 3.2, label.size = 0,
           fill = scales::alpha("white", 0.8), color = "grey20") +
  annotate("label", x =  Inf, y =  Inf, label = "High Growth / High Permits",
           hjust =  1.05, vjust = 1.05, size = 3.2, label.size = 0,
           fill = scales::alpha("white", 0.8), color = "grey20") +
  annotate("label", x = -Inf, y = -Inf, label = "Low Growth / Low Permits",
           hjust = -0.05, vjust = -0.05, size = 3.2, label.size = 0,
           fill = scales::alpha("white", 0.8), color = "grey20") +
  annotate("label", x =  Inf, y = -Inf, label = "High Growth / Low Permits",
           hjust =  1.05, vjust = -0.05, size = 3.2, label.size = 0,
           fill = scales::alpha("white", 0.8), color = "grey20")+


  # Labels for Top 20
  ggrepel::geom_label_repel(
    data = top_lab,
    aes(label = paste0(.data[[label_var]])),
    size = 3, label.size = 0.2, seed = 123,
    min.segment.length = 0, max.overlaps = Inf,
    fill = alpha("white", 0.8), color = "black"
  ) +

  # Scales
  scale_x_continuous(labels = percent_format(accuracy = 0.1),
                     name = "Population CAGR (5y)") +
  scale_y_continuous(name = "Housing Permits per 1,000 Residents (5y avg)") +
  scale_size_identity(guide = "none") +

  # Theme
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "Are Southern Metros Building Enough Housing?",
    subtitle = "Population growth vs. housing permits (5-year averages); Top 20 outlined & labeled",
    caption = "Notes: Points sized by population. Dashed lines mark regional medians."
  )

p_scatter
```
```{r Save Scatter chart}
# Best: set both plot and legend backgrounds to white
p_scatter <- p_scatter +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_growth_vs_housing_supply_scatter.png", p_scatter,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```

```{r Scatter with medians and quadrant labels}
# We need to improve the visual for quadrant labels that are stuck in the middle
df <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  filter(!is.na(pop_cagr_5y), !is.na(units_per_1k_5y_avg))

metric_x <- "pop_cagr_5y"
metric_y <- "units_per_1k_5y_avg"
rank_var <- "reg_rank_in_cbsa_type"

# Medians
x_med <- median(df[[metric_x]], na.rm = TRUE)
y_med <- median(df[[metric_y]], na.rm = TRUE)

# Top 20 for labels
top_lab <- df %>% arrange(.data[[rank_var]]) %>% slice_head(n = 20)

p_scatter <- ggplot(df, aes(x = .data[[metric_x]], y = .data[[metric_y]])) +
  geom_point(shape = 21, size = 3, stroke = 0.2, color = "grey35",
             fill = "#2C7FB8", alpha = 0.7) +

  # Outline Top 20
  geom_point(data = top_lab, shape = 21, size = 3.6, stroke = 1.0,
             color = "black", fill = NA) +

  # Median lines
  geom_vline(xintercept = x_med, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_hline(yintercept = y_med, linetype = "dashed", color = "grey50", linewidth = 0.4) +

  # Median value annotations
  annotate("label", x = x_med, y = min(df[[metric_y]], na.rm=TRUE),
           label = paste0("Median Pop CAGR: ", percent(x_med, 0.1)),
           vjust = 1.2, size = 3, label.size = 0, fill = alpha("white", 0.8)) +
  annotate("label", x = min(df[[metric_x]], na.rm=TRUE), y = y_med,
           label = paste0("Median Permits/1k: ", round(y_med,1)),
           hjust = -0.05, size = 3, label.size = 0, fill = alpha("white", 0.8)) +

  # Quadrant labels
  annotate("text", x = x_med*1.10, y = y_med*1.12, label = "High Growth / High Permits",
           hjust = 0, vjust = 0, size = 3.2, color = "grey30") +
  annotate("text", x = x_med*1.10, y = y_med*0.88, label = "High Growth / Low Permits",
           hjust = 0, vjust = 1, size = 3.2, color = "grey30") +
  annotate("text", x = x_med*0.90, y = y_med*1.12, label = "Low Growth / High Permits",
           hjust = 1, vjust = 0, size = 3.2, color = "grey30") +
  annotate("text", x = x_med*0.90, y = y_med*0.88, label = "Low Growth / Low Permits",
           hjust = 1, vjust = 1, size = 3.2, color = "grey30") +

  # Labels for Top 20
  ggrepel::geom_label_repel(
    data = top_lab,
    aes(label = paste0(NAME, ", ", primary_state)),
    size = 3, label.size = 0.2, seed = 123,
    min.segment.length = 0, max.overlaps = Inf,
    fill = alpha("white", 0.8), color = "black") +

  scale_x_continuous(labels = percent_format(accuracy = 0.1),
                     name = "Population CAGR (5y)") +
  scale_y_continuous(name = "Housing Permits per 1,000 Residents (5y avg)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "none",
        plot.title = element_text(face = "bold")) +
  labs(title = "Are Southern Metros Building Enough Housing?",
       subtitle = "Population growth vs. permitting; Top 20 outlined and labeled",
       caption = "Dashed lines mark Southern medians.")

p_scatter
```

Next, we will look at Vacancy vs Supply and Pop
```{r Vacancy Rate}
top20 <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area",
         !is.na(pop_cagr_5y), !is.na(VACANCY_RATE)) %>%
  arrange(desc(pop_cagr_5y)) %>%
  slice_head(n = 20) %>%
  mutate(metro_lbl = NAME)

vac_median <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  summarize(vac_med = median(VACANCY_RATE, na.rm = TRUE)) %>% pull(vac_med)

p_vac <- top20 %>%
  mutate(metro_lbl = factor(metro_lbl, levels = rev(metro_lbl))) %>%
  ggplot(aes(x = metro_lbl, y = VACANCY_RATE)) +
  geom_col(width = 0.7, fill = "#7F8DAA", color = "grey25") +
  geom_hline(yintercept = vac_median, linetype = "dashed", color = "grey40") +
  annotate("label", x = Inf, y = vac_median,
           label = paste0("Regional median: ", percent(vac_median, 0.1)),
           hjust = 1.02, vjust = -0.2, size = 3, label.size = 0,
           fill = alpha("white", 0.9)) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  coord_flip() +
  labs(
    title = "Vacancy Rates — Top 20 Fastest-Growing Southern Metros",
    subtitle = "Dashed line indicates regional median vacancy (metro areas only)",
    x = NULL, y = "Vacancy Rate (ACS)"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
p_vac
```
```{r Save Vacancy chart}
# Best: set both plot and legend backgrounds to white
p_vac <- p_vac +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_top_market_vacancy.png", p_vac,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```


# *Under Supplied Fast Growers*
Here we will create a Supply Gap metric and then visualize our top markets. We will calculate Supply Gap using the Z-Score method and also creating a flag of markets that are in the top growth quartile and below median in permits.

It can also be helpful to visualize top markets in terms of Z-Score if we break them out by the quadrant we used before. So we should save the classification from the scatter plot.

```{r Supply Gap DF}
# Base df (adjust names if needed)
df_supply_gap <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  filter(!is.na(pop_cagr_5y), !is.na(units_per_1k_5y_avg))

# Medians for quadrants (unweighted; simple & transparent)
x_med <- median(df_supply_gap$pop_cagr_5y, na.rm = TRUE)
y_med <- median(df_supply_gap$units_per_1k_5y_avg, na.rm = TRUE)

# Classification + supply gap
df_supply_gap_class <- df_supply_gap %>%
  mutate(
    # z-scores
    pop_z     = as.numeric(scale(pop_cagr_5y)),
    permits_z = as.numeric(scale(units_per_1k_5y_avg)),
    supply_gap = pop_z - permits_z,

    # “High growth, low supply” flags (Method A thresholds)
    pop_quartile = ntile(pop_cagr_5y, 4),
    below_perm_median = units_per_1k_5y_avg < y_med,
    high_growth_low_supply = pop_quartile == 4 & below_perm_median,

    # Quadrant classification (to match your scatter)
    quadrant_class = case_when(
      pop_cagr_5y >= x_med & units_per_1k_5y_avg >= y_med ~ "High Growth / High Permits",
      pop_cagr_5y >= x_med & units_per_1k_5y_avg <  y_med ~ "High Growth / Low Permits",
      pop_cagr_5y <  x_med & units_per_1k_5y_avg >= y_med ~ "Low Growth / High Permits",
      TRUE ~ "Low Growth / Low Permits"
    ),

    # Compact label
    metro_lbl = NAME
  )
```

```{r Supply Gap Visual}
# Palette to match (or echo) your scatter’s quadrant tints
quad_pal <- c(
  "High Growth / High Permits" = "#66bb6a",  # green-ish
  "High Growth / Low Permits"  = "#ffb74d",  # amber (undersupplied)
  "Low Growth / High Permits"  = "#42a5f5",  # blue
  "Low Growth / Low Permits"   = "#bdbdbd"   # grey
)

top_n <- 20

plot_df <- df_supply_gap_class %>%
  arrange(desc(supply_gap)) %>%
  slice_head(n = top_n) %>%
  mutate(metro_lbl = factor(metro_lbl, levels = rev(metro_lbl)))

p_supply_gap <- ggplot(plot_df, aes(x = metro_lbl, y = supply_gap, fill = quadrant_class)) +
  geom_col(width = 0.7, color = "grey25") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_text(aes(label = round(supply_gap, 2)), hjust = -0.06, size = 3) +
  scale_fill_manual(values = quad_pal, name = "Quadrant (from scatter)") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  labs(
    title = "Top Supply-Gap Metros in the South",
    subtitle = "Supply gap = z(Pop CAGR 5y) − z(Permits per 1k, 5y avg); color shows scatter-plot quadrant",
    x = NULL, y = "Supply Gap (standardized units)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

p_supply_gap
```

```{r Export Supply Gap chart}
# Best: set both plot and legend backgrounds to white
p_supply_gap <- p_supply_gap +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_top_supply_gap.png", p_supply_gap,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```

```{r Updated Vacancy Rate}
# Assume df_supply_gap_class already exists with quadrant_class assigned
# If not, you can re-use the classification logic from the supply gap step.

top20 <- df_supply_gap_class %>%
  filter(region == "South", cbsa_type == "Metro Area",
         !is.na(pop_cagr_5y), !is.na(VACANCY_RATE)) %>%
  arrange(desc(pop_cagr_5y)) %>%
  slice_head(n = 20) %>%
  mutate(metro_lbl = NAME)

vac_median <- df_supply_gap_class %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  summarize(vac_med = median(VACANCY_RATE, na.rm = TRUE)) %>% pull(vac_med)

p_vac <- top20 %>%
  mutate(metro_lbl = factor(metro_lbl, levels = rev(metro_lbl))) %>%
  ggplot(aes(x = metro_lbl, y = VACANCY_RATE, fill = quadrant_class)) +
  geom_col(width = 0.7, color = "grey25") +
  geom_hline(yintercept = vac_median, linetype = "dashed", color = "grey40") +
  annotate("label", x = Inf, y = vac_median,
           label = paste0("Regional median: ", percent(vac_median, 0.1)),
           hjust = 1.02, vjust = -0.2, size = 3, label.size = 0,
           fill = alpha("white", 0.9)) +
  scale_fill_manual(values = quad_pal, name = "Quadrant (from scatter)") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  coord_flip() +
  labs(
    title = "Vacancy Rates — Top 20 Fastest-Growing Southern Metros",
    subtitle = "Color shows scatter-plot quadrant; dashed line = regional median vacancy",
    x = NULL, y = "Vacancy Rate (ACS)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_vac
```

```{r Export Vacancy Updated chart}
# Best: set both plot and legend backgrounds to white
p_vac <- p_vac +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_top_market_vacancy.png", p_vac,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```


```{r Simple Supply Gap}
df_gap <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  filter(!is.na(pop_cagr_5y), !is.na(units_per_1k_5y_avg)) %>%
  mutate(
    pop_quartile = ntile(pop_cagr_5y, 4),
    permits_median = units_per_1k_5y_avg < median(units_per_1k_5y_avg, na.rm = TRUE),
    undersupplied_flag = pop_quartile == 4 & permits_median
  )

# Pull the subset for the article
undersupplied_markets <- df_gap %>%
  filter(undersupplied_flag) %>%
  arrange(desc(pop_cagr_5y)) %>%
  select(NAME, primary_state, pop_cagr_5y, units_per_1k_5y_avg,
         gdp_real_cagr_5y, unemployment_rate, TOTAL_POP)
```

```{r Continious Supply Gap}
df_gap <- df_gap %>%
  mutate(
    pop_z = scale(pop_cagr_5y)[,1],
    permits_z = scale(units_per_1k_5y_avg)[,1],
    supply_gap = pop_z - permits_z
  )

top_supply_gaps <- df_gap %>%
  arrange(desc(supply_gap)) %>%
  slice_head(n = 10) %>%
  select(NAME, primary_state, pop_cagr_5y, units_per_1k_5y_avg,
         supply_gap, gdp_real_cagr_5y, unemployment_rate, TOTAL_POP)
```

```{r Supply Gap Chart}
df_gap <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  filter(!is.na(pop_cagr_5y), !is.na(units_per_1k_5y_avg))

# Thresholds
perm_med <- median(df_gap$units_per_1k_5y_avg, na.rm = TRUE)

df_gap <- df_gap %>%
  mutate(
    pop_quartile   = ntile(pop_cagr_5y, 4),
    undersupplied  = pop_quartile == 4 & units_per_1k_5y_avg < perm_med,
    pop_z          = as.numeric(scale(pop_cagr_5y)),
    permits_z      = as.numeric(scale(units_per_1k_5y_avg)),
    supply_gap     = pop_z - permits_z,
    metro_lbl      = str_trunc(paste0(NAME, ", ", primary_state), 38)
  )

# Pick the story you want to show:
# A) Only the undersupplied fast-growers, ranked by supply_gap
plot_df <- df_gap %>%
  filter(undersupplied) %>%
  arrange(desc(supply_gap)) %>%
  slice_head(n = 10)   # change to 15/20 if you like

# B) Or: top N overall by supply_gap (comment A, uncomment B)
# plot_df <- df_gap %>% arrange(desc(supply_gap)) %>% slice_head(n = 10)

# Bar chart
p_gap <- plot_df %>%
  mutate(metro_lbl = factor(metro_lbl, levels = rev(metro_lbl))) %>%
  ggplot(aes(x = metro_lbl, y = supply_gap)) +
  geom_col(width = 0.7, fill = "#2C7FB8", color = "grey25") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_text(aes(label = round(supply_gap, 2)), hjust = -0.05, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  labs(
    title = "Undersupplied Fast-Growers — Supply Gap",
    subtitle = "Supply gap = z(Pop CAGR 5y) − z(Permits per 1k, 5y avg); Top markets where growth outpaces permitting",
    x = NULL, y = "Supply Gap (standardized units)"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
p_gap
```



# *Demographics*
Education 
Age
Labor Force Participation

```{r Median Age Comparison}
# Split metros into top 20 and bottom 20 by growth
df_age <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area", !is.na(pop_cagr_5y), !is.na(MEDIAN_AGE)) %>%
  mutate(rank_growth = min_rank(desc(pop_cagr_5y))) %>%
  mutate(group = case_when(
    rank_growth <= 20 ~ "Top 20 by Growth",
    rank_growth > (n() - 20) ~ "Bottom 20 by Growth",
    TRUE ~ "Middle"
  ))

p_age <- df_age %>%
  filter(group %in% c("Top 20 by Growth", "Bottom 20 by Growth")) %>%
  ggplot(aes(x = group, y = MEDIAN_AGE, fill = group)) +
  geom_boxplot(alpha = 0.7, width = 0.6) +
  scale_fill_manual(values = c("Top 20 by Growth" = "#2C7FB8", "Bottom 20 by Growth" = "#bdbdbd")) +
  labs(
    title = "Median Age in Southern Metros",
    subtitle = "Comparing fastest- and slowest-growing metros",
    x = NULL, y = "Median Age (ACS)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
p_age
```
Interesting insight that the median ages seem very close but the distributions of the ages for the markets are very different.

```{r Export Age chart}
# Best: set both plot and legend backgrounds to white
p_age <- p_age +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_top_vs_bottom_age.png", p_age,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```


```{r Education Scatter Plot}
df_edu <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area", !is.na(pop_cagr_5y), !is.na(HIGHER_ED_RATE)) %>%
  mutate(HIGHER_ED_RATE = HIGHER_ED_RATE / 100)

p_edu <- ggplot(df_edu, aes(x = pop_cagr_5y, y = HIGHER_ED_RATE, label = NAME)) +
  geom_point(aes(size = TOTAL_POP, color = division), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "grey40", linetype = "dashed") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Population Growth vs Educational Attainment",
    subtitle = "% of population with Bachelor's+ (ACS, 2023)",
    x = "Population CAGR (5y)", y = "% with Bachelor’s Degree or Higher",
    color = "State", size = "Population"
  ) +
  theme_minimal(base_size = 11)
p_edu
```
```{r Export Education Chart}
# Best: set both plot and legend backgrounds to white
p_edu <- p_edu +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_pop_growth_vs_education.png", p_edu,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```


```{r Labor Force Participation}
df_lfpr <- acs_silver %>%   # assuming you have yearly ACS silver with "in_labor_force" and "eligible_labor_force"
  mutate(LFP_RATE = labor_force_participation_rate / 100,
         GEOID = as.character(GEOID))

# Rank by growth
growth_ranks <- analysis_final_table %>%
  filter(cbsa_type == "Metro Area") %>%
  select(GEOID, pop_cagr_5y) %>%
  mutate(rank_growth = min_rank(desc(pop_cagr_5y))) %>%
  mutate(group = case_when(
    rank_growth <= 20 ~ "Top 20 by Growth",
    rank_growth > (n() - 20) ~ "Bottom 20 by Growth",
    TRUE ~ "Middle"
  ))

df_lfpr <- df_lfpr %>%
  right_join(growth_ranks, by = "GEOID") %>%
  filter(year %in% c(2013, 2023))

p_lfpr <- df_lfpr %>%
  filter(group %in% c("Top 20 by Growth", "Bottom 20 by Growth")) %>%
  ggplot(aes(x = year, y = LFP_RATE, group = NAME, color = group)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Labor Force Participation, 2013–2023",
    subtitle = "Comparing fastest- and slowest-growing Southern metros",
    x = NULL, y = "Labor Force Participation Rate",
    color = NULL
  ) +
  theme_minimal(base_size = 11)
p_lfpr
```

```{r Export LFPR Chart}
# Best: set both plot and legend backgrounds to white
p_lfpr <- p_lfpr +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_lfpr_change.png", p_lfpr,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```

# *Economics*
How does Population Growth compare to GDP Growth?
Bubble Plot: X = pop CAGR, Y = GDP CAGR, bubble size = income level.
Slope Chart: Pop growth vs income growth (metro by metro).

```{r Pop Growth vs GDP Growth}
econ_df <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  filter(!is.na(pop_cagr_5y), !is.na(gdp_real_cagr_5y), !is.na(MEDIAN_INCOME))

x_med <- median(econ_df$pop_cagr_5y, na.rm = TRUE)
y_med <- median(econ_df$gdp_real_cagr_5y, na.rm = TRUE)

# Select Top 20 by population growth
top_lab <- econ_df %>%
  arrange(desc(pop_cagr_5y)) %>%
  slice_head(n = 20)

# GDP vs Pop Bubble with labels
p_gdp_v_pop_bubble <- ggplot(econ_df,
                   aes(x = pop_cagr_5y, y = gdp_real_cagr_5y,
                       size = TOTAL_POP, color = division)) +
  geom_point(alpha = 0.7, stroke = 0.2) +
  # regression trend (this was missing before)
  geom_smooth(method = "lm", se = FALSE, color = "grey35",
              linetype = "longdash", linewidth = 0.8,
              inherit.aes = FALSE, show.legend = FALSE,
              mapping = aes(x = pop_cagr_5y, y = gdp_real_cagr_5y)) +
  geom_point(data = top_lab, shape = 21, size = 4, stroke = 1.0,
             color = "black", fill = NA, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_lab,
    aes(label = NAME),
    size = 3, seed = 123,
    box.padding = 0.3, max.overlaps = Inf, show.legend = FALSE
  ) +
  geom_vline(xintercept = x_med, linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = y_med, linetype = "dashed", color = "grey60") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                     name = "Total Population CAGR (5y)") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     name = "Real GDP CAGR (5y)") +
  scale_size_continuous(name = "Total Population", labels = scales::label_comma(),
                        range = c(1.8, 9)) +
  guides(size = guide_legend(override.aes = list(alpha = 0.9, color = "grey50", stroke = 0))) +

  labs(title = "Are Southern Metros Converting Population Growth into Higher GDP?",
       subtitle = "Top 20 growth metros outlined and labeled",
       color = "Division") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
p_gdp_v_pop_bubble
```

```{r Save GDP Growth}
# Best: set both plot and legend backgrounds to white
p_gdp_v_pop_bubble <- p_gdp_v_pop_bubble +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_gdp_growth_vs_population.png", p_gdp_v_pop_bubble,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```

```{r Income vs Population}
econ_df <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  filter(!is.na(pop_cagr_5y), !is.na(income_cagr_5y), !is.na(MEDIAN_INCOME))

x_med <- median(econ_df$pop_cagr_5y, na.rm = TRUE)
y_med <- median(econ_df$income_cagr_5y, na.rm = TRUE)

# Select Top 20 by population growth
top_lab <- econ_df %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  arrange(desc(pop_cagr_5y)) %>%
  slice_head(n = 20)

# Population vs Income Bubble with labels
p_inc_v_pop_bubble <- ggplot(econ_df,
                   aes(x = pop_cagr_5y, y = income_cagr_5y,
                       size = TOTAL_POP, color = division)) +
  geom_point(alpha = 0.7, stroke = 0.2) +
  # regression trend (this was missing before)
  geom_smooth(method = "lm", se = FALSE, color = "grey35",
              linetype = "longdash", linewidth = 0.8,
              inherit.aes = FALSE, show.legend = FALSE,
              mapping = aes(x = pop_cagr_5y, y = income_cagr_5y)) +
  geom_point(data = top_lab, shape = 21, size = 4, stroke = 1.0,
             color = "black", fill = NA, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_lab,
    aes(label = NAME),
    size = 3, seed = 123,
    box.padding = 0.3, max.overlaps = Inf, show.legend = FALSE
  ) +
  geom_vline(xintercept = x_med, linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = y_med, linetype = "dashed", color = "grey60") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                     name = "Total Population CAGR (5y)") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     name = "Median Income CAGR (5y)") +
  scale_size_continuous(name = "Total Population", labels = scales::label_comma(),
                        range = c(1.8, 9)) +
  guides(size = guide_legend(override.aes = list(alpha = 0.9, color = "grey50", stroke = 0))) +

  labs(title = "Are Southern Metros Converting Population Growth into Higher Income?",
       subtitle = "Top 20 growth metros outlined and labeled",
       color = "Division") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
p_inc_v_pop_bubble
```

This is an interesting relationship. Seems like a few outliers are really skewing the line to the right.

```{r Save Income Growth}
# Best: set both plot and legend backgrounds to white
p_inc_v_pop_bubble <- p_inc_v_pop_bubble +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_income_growth_vs_population.png", p_inc_v_pop_bubble,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```


```{r Income vs GDP}
# Base df for this chart
econ_df <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  filter(!is.na(gdp_real_cagr_5y), !is.na(income_cagr_5y),
         !is.na(MEDIAN_INCOME), !is.na(TOTAL_POP))

x_med <- median(econ_df$gdp_real_cagr_5y, na.rm = TRUE)
y_med <- median(econ_df$income_cagr_5y, na.rm = TRUE)

# Select Top 20 by population growth
top_lab <- econ_df %>%
  arrange(desc(pop_cagr_5y)) %>%
  slice_head(n = 20)

# GDP vs Income Bubble with labels
p_inc_v_gdp_bubble <- ggplot(econ_df,
                   aes(x = gdp_real_cagr_5y, y = income_cagr_5y,
                       size = TOTAL_POP, color = division)) +
  geom_point(alpha = 0.7, stroke = 0.2) +
  # regression trend (this was missing before)
  geom_smooth(method = "lm", se = FALSE, color = "grey35",
              linetype = "longdash", linewidth = 0.8,
              inherit.aes = FALSE, show.legend = FALSE,
              mapping = aes(x = gdp_real_cagr_5y, y = income_cagr_5y)) +
  geom_point(data = top_lab, shape = 21, size = 4, stroke = 1.0,
             color = "black", fill = NA, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_lab,
    aes(label = NAME),
    size = 3, seed = 123,
    box.padding = 0.3, max.overlaps = Inf, show.legend = FALSE
  ) +
  geom_vline(xintercept = x_med, linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = y_med, linetype = "dashed", color = "grey60") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                     name = "Real GDP CAGR (5y)") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     name = "Median Income CAGR (5y)") +
  scale_size_continuous(name = "Total Population", labels = scales::label_comma(),
                        range = c(1.8, 9)) +
  guides(size = guide_legend(override.aes = list(alpha = 0.9, color = "grey50", stroke = 0))) +

  labs(title = "Are Southern Metros Converting GDP Growth into Higher Income?",
       subtitle = "Top 20 growth metros outlined and labeled",
       color = "Division") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
p_inc_v_gdp_bubble
```

```{r Save Income vs GDP}
# Best: set both plot and legend backgrounds to white
p_inc_v_gdp_bubble <- p_inc_v_gdp_bubble +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_gdp_vs_income_growth.png", p_inc_v_gdp_bubble,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```

```{r Fundamentals Negative Gap}
econ_gap <- analysis_final_table %>%
  filter(region == "South", cbsa_type == "Metro Area") %>%
  transmute(
    GEOID, NAME, primary_state, TOTAL_POP,
    pop = pop_cagr_5y,
    gdp = gdp_real_cagr_5y,
    inc = income_cagr_5y,
    unemp = unemployment_rate
  ) %>%
  filter(!is.na(pop), !is.na(gdp), !is.na(inc)) %>%
  mutate(
    z_pop = as.numeric(scale(pop)),
    z_gdp = as.numeric(scale(gdp)),
    z_inc = as.numeric(scale(inc)),
    fundamentals_gap = (z_gdp + z_inc)/2 - z_pop
  )

# Rank top/bottom outliers
top_pos_gap <- econ_gap %>% arrange(desc(fundamentals_gap)) %>% slice_head(n = 10)
top_neg_gap <- econ_gap %>% arrange(fundamentals_gap) %>% slice_head(n = 10)

# Visual: ordered bar chart of negative gaps (population > economics)
gap_neg_plot <- econ_gap %>%
  arrange(fundamentals_gap) %>%
  slice_head(n = 15) %>%
  mutate(lbl = factor(NAME,
                      levels = NAME)) %>%
  ggplot(aes(x = lbl, y = fundamentals_gap)) +
  geom_col(fill = "#ffb74d", color = "grey25") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  coord_flip() +
  labs(title = "Where Population Is Outrunning the Economics",
       subtitle = "Fundamentals gap = avg z(GDP growth, income growth) − z(pop growth)",
       x = NULL, y = "Fundamentals Gap (z-score units)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
gap_neg_plot
```
```{r Save Fundamentals Negative}
# Best: set both plot and legend backgrounds to white
gap_neg_plot <- gap_neg_plot +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_fundamental_gap_negative.png", gap_neg_plot,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```


```{r Fundamentals Positive}
# Visual: ordered bar chart of negative gaps (population > economics)
gap_pos_plot <- econ_gap %>%
  arrange(desc(fundamentals_gap)) %>%
  slice_head(n = 15) %>%
  mutate(lbl = factor(NAME,
                      levels = NAME)) %>%
  ggplot(aes(x = lbl, y = fundamentals_gap)) +
  geom_col(fill = "#ffb74d", color = "grey25") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  coord_flip() +
  labs(title = "Where Economics Is Outrunning the Population",
       subtitle = "Fundamentals gap = avg z(GDP growth, income growth) − z(pop growth)",
       x = NULL, y = "Fundamentals Gap (z-score units)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
gap_pos_plot
```

```{r Save Fundamentals Positive}
# Best: set both plot and legend backgrounds to white
gap_pos_plot <- gap_pos_plot +
  theme(
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key        = element_rect(fill = "white", color = NA)
  )

ggsave("/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_fundamental_gap_positive.png", gap_pos_plot,
       width = 12, height = 8, dpi = 300,
       bg = "white",
  device = ragg::agg_png)
```

```{r Export Econ DF}
# Add Fundamental Gap to Final Table to export
export_econ_df <- econ_df %>%
  left_join(econ_gap %>% select(GEOID, fundamentals_gap), by = "GEOID")

write_csv(export_econ_df, "/Users/danberle/Documents/projects/demographic_analysis/outputs/cbsa_econ_df.csv")
```


# *Final Recommendations*
A summary table with the top 10 rankings and their core KPIs



*Risk*
We can add this later




#Appendix
Questions:
- Why are some CBSAs missing from our growth data frame. For example: "17410" - Cleveland 
- Let's go back and create rankings for Metro vs Micro Areas so we can split this better


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

