---
title: "Retail Opportunity Finder MVP"
subtitle: "Integration Scaffold (Artifact-Driven)"
format:
  html:
    page-layout: full
    toc: true
    toc-location: left
    toc-title: "Contents"
    toc-expand: true
    toc-depth: 3
    number-sections: true
    grid:
      sidebar-width: 260px
      body-width: 1200px
      margin-width: 260px
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: f2-load-artifacts
#| include: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(readr)
  library(ggplot2)
  library(gt)
  library(sf)
  library(DT)
  library(htmltools)
})

resolve_project_root <- function() {
  candidates <- c(".", "..", "../..", "../../..", "../../../..", "../../../../..")
  for (cand in candidates) {
    root <- normalizePath(cand, winslash = "/", mustWork = FALSE)
    marker <- file.path(root, "notebooks/retail_opportunity_finder/sections")
    if (dir.exists(marker)) return(root)
  }
  stop("Unable to resolve project root from current render working directory.", call. = FALSE)
}

project_root <- resolve_project_root()
source(file.path(project_root, "notebooks/retail_opportunity_finder/sections/_shared/bootstrap.R"))
initialize_section_runtime()

resolve_artifact_path <- function(path) {
  if (file.exists(path)) return(path)
  rooted <- file.path(project_root, path)
  if (file.exists(rooted)) return(rooted)
  path
}

resolve_section_artifact <- function(section_id, artifact_name, subdir = NULL, ext = "rds") {
  read_artifact_path(section_id, artifact_name, key = ACTIVE_MARKET_KEY, subdir = subdir, ext = ext)
}

read_artifact <- function(path, label) {
  resolved <- resolve_artifact_path(path)
  if (!file.exists(resolved)) {
    stop(paste0("Missing required artifact [", label, "]: ", path), call. = FALSE)
  }
  readRDS(resolved)
}

artifact_manifest <- tribble(
  ~alias, ~path,
  "section01_run_metadata", resolve_section_artifact("01_setup", "section_01_run_metadata"),
  "section01_foundation", resolve_section_artifact("01_setup", "section_01_foundation"),

  "section02_visual_objects", resolve_section_artifact("02_market_overview", "section_02_visual_objects"),
  "section02_validation_report", resolve_section_artifact("02_market_overview", "section_02_validation_report"),

  "section03_funnel_counts", resolve_section_artifact("03_eligibility_scoring", "section_03_funnel_counts"),
  "section03_scored_tracts", resolve_section_artifact("03_eligibility_scoring", "section_03_scored_tracts"),
  "section03_top_tracts", resolve_section_artifact("03_eligibility_scoring", "section_03_top_tracts"),
  "section03_visual_objects", resolve_section_artifact("03_eligibility_scoring", "section_03_visual_objects"),
  "section03_validation_report", resolve_section_artifact("03_eligibility_scoring", "section_03_validation_report"),

  "section04_cluster_zone_summary", resolve_section_artifact("04_zones", "section_04_cluster_zone_summary"),
  "section04_cluster_visual_objects", resolve_section_artifact("04_zones", "section_04_cluster_visual_objects"),
  "section04_cluster_validation_report", resolve_section_artifact("04_zones", "section_04_cluster_validation_report"),

  "section05_zone_overlay_cluster", resolve_section_artifact("05_parcels", "section_05_zone_overlay_cluster"),
  "section05_parcel_shortlist_cluster", resolve_section_artifact("05_parcels", "section_05_parcel_shortlist_cluster"),
  "section05_visual_objects", resolve_section_artifact("05_parcels", "section_05_visual_objects"),
  "section05_validation_report", resolve_section_artifact("05_parcels", "section_05_validation_report"),

  "section06_conclusion_payload", resolve_section_artifact("06_conclusion_appendix", "section_06_conclusion_payload"),
  "section06_appendix_payload", resolve_section_artifact("06_conclusion_appendix", "section_06_appendix_payload"),
  "section06_visual_objects", resolve_section_artifact("06_conclusion_appendix", "section_06_visual_objects"),
  "section06_validation_report", resolve_section_artifact("06_conclusion_appendix", "section_06_validation_report")
)

artifacts <- lapply(seq_len(nrow(artifact_manifest)), function(i) {
  read_artifact(artifact_manifest$path[[i]], artifact_manifest$alias[[i]])
})
names(artifacts) <- artifact_manifest$alias

# Section-level aliases for clean downstream chunks
run_meta <- artifacts$section01_run_metadata
foundation <- artifacts$section01_foundation
market_profile <- foundation$market_profile
market_labels <- market_profile$labels
target_market_name <- market_labels$market_name
target_flag_label <- market_labels$target_flag
benchmark_region_label <- market_profile$benchmark_region_label

s02_visuals <- artifacts$section02_visual_objects
s03_visuals <- artifacts$section03_visual_objects
s04_cluster_visuals <- artifacts$section04_cluster_visual_objects
s05_visuals <- artifacts$section05_visual_objects
s06_visuals <- artifacts$section06_visual_objects

section_02_report <- artifacts$section02_validation_report
section_03_report <- artifacts$section03_validation_report
section_04_cluster_report <- artifacts$section04_cluster_validation_report
section_05_report <- artifacts$section05_validation_report
section_06_report <- artifacts$section06_validation_report

# Defensive validation guards
pass_02 <- all(vapply(section_02_report$checks, function(x) isTRUE(x$pass), logical(1))) &&
  all(unlist(section_02_report$logic_checks))
pass_03 <- all(vapply(section_03_report$checks[1:5], function(x) isTRUE(x$pass), logical(1))) &&
  isTRUE(section_03_report$checks$geometry_check$pass) &&
  all(unlist(section_03_report$logic_checks))

if (!isTRUE(pass_02)) stop("Section 02 validation guard failed.", call. = FALSE)
if (!isTRUE(pass_03)) stop("Section 03 validation guard failed.", call. = FALSE)
if (!isTRUE(section_04_cluster_report$pass)) stop("Section 04 cluster validation guard failed.", call. = FALSE)
if (!isTRUE(section_05_report$pass)) stop("Section 05 validation guard failed.", call. = FALSE)
if (!isTRUE(section_06_report$pass)) stop("Section 06 validation guard failed.", call. = FALSE)

section_04_warnings <- if (!is.null(section_04_cluster_report$warnings)) section_04_cluster_report$warnings else character()
section_05_warnings <- if (!is.null(section_05_report$warnings)) section_05_report$warnings else character()
section_05_coverage <- section_05_report$coverage_metrics

required_visual_keys <- list(
  s02_visuals = c("tiles_layout", "peer_gt", "benchmark_gt", "pop_trend_plot", "distribution_plot"),
  s03_visuals = c("eligible_map_plot", "score_hist_plot", "component_mix_plot", "top_tracts_gt"),
  s04_cluster_visuals = c("cluster_zone_map_plot", "cluster_summary_gt"),
  s05_visuals = c("market_parcel_context_map", "cluster_parcel_overlay_map", "overlay_map_cluster", "shortlist_map_cluster", "shortlist_table_cluster"),
  s06_visuals = c("conclusion_summary_table", "shortlist_summary_table", "qa_summary_table", "assumptions_caveats_table", "recommendations_table")
)

visual_lists <- list(
  s02_visuals = s02_visuals,
  s03_visuals = s03_visuals,
  s04_cluster_visuals = s04_cluster_visuals,
  s05_visuals = s05_visuals,
  s06_visuals = s06_visuals
)

for (nm in names(required_visual_keys)) {
  missing_keys <- setdiff(required_visual_keys[[nm]], names(visual_lists[[nm]]))
  if (length(missing_keys) > 0) {
    stop(paste0("Missing visual object keys in ", nm, ": ", paste(missing_keys, collapse = ", ")), call. = FALSE)
  }
}

runtime_manifest <- artifact_manifest %>% mutate(exists = TRUE)

integration_outputs_dir <- file.path(project_root, "notebooks/retail_opportunity_finder/integration/outputs")
if (!dir.exists(integration_outputs_dir)) {
  dir.create(integration_outputs_dir, recursive = TRUE)
}
readr::write_csv(runtime_manifest, file.path(integration_outputs_dir, "phase_f2_runtime_manifest.csv"))
```

```{r}
#| label: f2-status
#| include: false

status_tbl <- tibble::tibble(
  checkpoint = c(
    "Artifact manifest loaded",
    "Section validation guards",
    "Visual object key guards",
    "Runtime mode"
  ),
  status = c("PASS", "PASS", "PASS", "ARTIFACT-ONLY"),
  detail = c(
    paste0(nrow(runtime_manifest), " artifacts loaded via readRDS"),
    "Sections 02-06 validation checks passed",
    "All required visual object names present",
    "No section build/visual/check scripts executed in notebook"
  )
)

status_tbl %>% gt::gt()
```

# Setup and Method

This report helps us find where retail corridor opportunity is most promising in the `r target_market_name`.
We move from broad market context to tract-level screening, then prioritize areas for deeper site review.

```{r}
#| label: s01-scaffold-status-hidden
#| include: false

status_tbl %>% gt::gt()
```

The output is practical:

- A shortlist of high-opportunity geographies
- A clear explanation of why those places rank higher
- A repeatable framework we can apply in other metros

## How to Read This Notebook

Think of the process in three steps:

1. **Market context:**
We start with a quick check of `r target_flag_label`'s growth, housing pressure, and peer position.
2. **Tract screening and ranking:**
We evaluate tracts using a consistent scorecard to identify where demand momentum and development headroom are strongest.
3. **Action-oriented prioritization:**
We carry top-performing areas into zone and parcel analysis to support field diligence and underwriting.

## Key Assumptions (Business View)

This model is designed for screening, not final investment decisions.
It is meant to narrow the market to better starting points for human review.

Important assumptions:

- We compare tracts within the same metro for fairness.
- We prioritize growth and housing pipeline signals.
- We treat pricing pressure as a guardrail, not the only decision factor.
- Final decisions still require local market diligence (tenants, zoning, access, competition).

<details>
<summary><strong>Method Transparency (for technical readers)</strong></summary>

- Data sources include ACS, BPS, tract geometry, and parcel attributes from validated section artifacts.
- Metrics are standardized within the target metro to support apples-to-apples tract comparison.
- The score combines weighted components aligned to the strategy thesis: growth, housing pipeline, development headroom, price pressure, and commute context.
- Gates are applied to enforce strategy fit and reduce false positives.
- Validation reports are generated by section and surfaced in the appendix.
- This is a screening model. Site-level diligence and underwriting are still required.

</details>

```{r}
#| label: s01-run-metadata

run_meta_tbl <- tibble::tibble(
  field = c("Run timestamp", "R version", "Git hash", "Market key", "Market name", "Target CBSA", "Target year"),
  value = c(
    run_meta$run_timestamp,
    run_meta$r_version,
    run_meta$git_hash,
    run_meta$market_key,
    run_meta$market_name,
    run_meta$target_cbsa,
    as.character(run_meta$target_year)
  )
)

run_meta_tbl %>%
  gt::gt() %>%
  gt::tab_header(title = "Run Metadata")
```

```{r}
#| label: s01-model-params
#| include: false

cluster_top_share_display <- if (!is.null(foundation$model_params$cluster_top_share)) {
  scales::percent(foundation$model_params$cluster_top_share, accuracy = 1)
} else {
  "25%"
}

model_params_tbl <- tibble::tibble(
  parameter = c(
    "Minimum growth gate percentile",
    "Maximum price proxy percentile",
    "Maximum density percentile",
    "Cluster cutoff (top share of scored tracts)",
    "Top N tracts for ranking output",
    "Score weights (growth/units/headroom/price/commute/income)"
  ),
  value = c(
    as.character(foundation$model_params$min_growth_percentile),
    as.character(foundation$model_params$price_proxy_pctl_max),
    as.character(foundation$model_params$max_density_percentile),
    cluster_top_share_display,
    as.character(foundation$model_params$top_n_tracts),
    paste(as.character(foundation$model_params$weights), collapse = " / ")
  )
)

model_params_tbl %>%
  gt::gt() %>%
  gt::tab_header(title = "Model Configuration Snapshot")
```

```{r}
#| label: s01-kpi-dictionary

foundation$kpi_dictionary %>%
  gt::gt() %>%
  gt::tab_header(title = "KPI Dictionary (V1)")
```

# Market Overview

Before drilling into tracts, we confirm whether `r target_flag_label` has the type of market conditions where retail corridor expansion is plausible.

What we look for:

- Population growth momentum
- Housing supply pipeline
- Relative affordability pressure
- Position versus peers and broader benchmarks

What this means:

- If metro-level setup is weak, tract-level targeting is less reliable.
- If metro-level signals are constructive, tract-level analysis becomes more actionable.

Unless otherwise noted, population growth is shown as share/percent, housing cost metrics are in USD, and supply is units per 1,000 residents.

```{r}
#| label: s02-market-context-map
#| fig-height: 6
#| fig-width: 9
#| fig-cap: "Market context map with tract growth signal and contextual boundaries/transport layers when available."

s02_visuals$market_context_map_plot
```

```{r}
#| label: s02-kpi-tiles

s02_visuals$tiles_layout
```

```{r}
#| label: s02-peer-table

s02_visuals$peer_gt
```

```{r}
#| label: s02-benchmark-table

s02_visuals$benchmark_gt
```

```{r}
#| label: s02-pop-trend
#| fig-height: 5
#| fig-width: 10
#| fig-cap: "Indexed population trend (baseline = 100). Source: Section 02 artifacts derived from `cbsa_features`."

s02_visuals$pop_trend_plot
```

```{r}
#| label: s02-distribution
#| fig-height: 6
#| fig-width: 11
#| fig-cap: "Target market position vs all metro distributions by key metrics. Source: Section 02 artifacts."

s02_visuals$distribution_plot
```

# Section 3 Eligibility and Scoring

This section ranks every tract first, then applies strategy gates so we can focus on the strongest candidates.

**What we do:**

- Build one tract score across all tracts using growth, supply, headroom, price pressure, commute exposure, and income.
- Use the top `25%` of scored tracts as the clustering cutoff universe.
- Keep gates as secondary diagnostics for strategy-fit storytelling.
- Surface the top tracts with plain-language why tags.

**What this means:**

- Higher-ranked tracts are stronger candidates for zone and parcel follow-up.
- Lower-ranked tracts are deprioritized, not permanently excluded from future review.
- The model provides a transparent starting point, not a final verdict.

```{r}
#| label: s03-scorecard

scorecard_tbl <- tibble::tibble(
  variable = c(
    "Population growth (3y)",
    "Units per 1k (3y avg)",
    "Population density",
    "Price proxy percentile",
    "Commute intensity",
    "Median household income"
  ),
  weight = c(
    foundation$model_params$weights["growth"],
    foundation$model_params$weights["units"],
    foundation$model_params$weights["headroom"],
    foundation$model_params$weights["price"],
    foundation$model_params$weights["commute"],
    foundation$model_params$weights["income"]
  ),
  effect = c("Positive", "Positive", "Negative", "Negative", "Positive", "Positive")
) %>%
  mutate(weight_pct = scales::percent(weight, accuracy = 1)) %>%
  select(variable, weight_pct, effect) %>%
  gt::gt() %>%
  gt::tab_header(title = "Scoring Variables, Weights, and Direction") %>%
  gt::cols_label(
    variable = "Variable",
    weight_pct = "Weight",
    effect = "Effect on score"
  )

scorecard_tbl
```

Clustering cutoff uses the top `25%` of tracts by `tract_score`, where each component is standardized within the metro.

```{r}
#| label: s03-eligible-map
#| fig-height: 6
#| fig-width: 9
#| fig-cap: "Tracts shaded by composite score across the full metro footprint, with county/context overlays. Source: Section 03 + Section 02 context artifacts."

s03_visuals$eligible_map_plot
```

```{r}
#| label: s03-component-mix
#| fig-height: 5
#| fig-width: 9
#| fig-cap: "Average contribution to score among top shortlisted tracts (eligible)."

s03_visuals$component_mix_plot
```

Method note: This chart averages each `contrib_*` component across top shortlisted tracts. Positive values increase score; negative values indicate that component sits below metro baseline even for top-ranked tracts.

```{r}
#| label: s03-top-tracts

htmltools::div(
  style = "max-height: 520px; overflow-y: auto; border: 1px solid #d9d9d9; border-radius: 4px; padding: 8px;",
  s03_visuals$top_tracts_gt
)
```

# Zone Systems

This section turns the Section 3 top-scoring tract set into operating zones for field review and corridor prioritization.

How zones are built:

1. Start from the top `25%` of tracts by Section 3 tract score (cluster seed universe).
2. Project tract centroids and connect nearby tracts using a fixed distance threshold.
3. Form connected components, then assign small/noise groups to the nearest core cluster.
4. Dissolve tract groups into cluster zone polygons and compute zone summary metrics.

Cluster summary for current run:

- Cluster zones: `r nrow(artifacts$section04_cluster_zone_summary)`
- Cluster seed tracts: `r nrow(readRDS(resolve_section_artifact("04_zones", "section_04_zone_input_candidates")))`

```{r}
#| label: s04-caveat-note

if (length(section_04_warnings) > 0) {
  caveat_tbl <- tibble::tibble(
    caveat = section_04_warnings
  )
  caveat_tbl %>%
    gt::gt() %>%
    gt::tab_header(title = "Section 04 Caveat")
}
```

```{r}
#| label: s04-cluster-map
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Cluster zone map generated from top-scoring tract seeds. Source: Section 04 cluster artifacts."

s04_cluster_visuals$cluster_zone_map_plot
```

```{r}
#| label: s04-cluster-summary

s04_cluster_visuals$cluster_summary_gt
```

Operational read:

- Tighter clusters improve leasing and field-routing focus.
- Zone summaries provide a direct bridge into parcel-level overlay in Section 5.

# Retail Overlay and Shortlist

Section 05 translates high-opportunity zones into a practical parcel shortlist for broker and field review.

How the shortlist score works:
- `Zone quality` (50%): stronger cluster zones score higher based on upstream tract performance.
- `Local retail context` (25%): parcels in tracts with stronger existing retail intensity score higher.
- `Parcel characteristics` (25%): favors larger parcels, lower value-per-square-foot, and more recent sales activity.

Value and sales logic:
- `JV (Just Value)` is the primary value signal.
- We compute value intensity as `JV / land square feet` and winsorize extremes for stability.
- `SALE_PRC1` is retained for sales context and recency (via `last_sale_date`), not as the primary value field.

Current run snapshot:

- Cluster shortlist candidates: `r nrow(artifacts$section05_parcel_shortlist_cluster)`
- Covered cluster zones: `r dplyr::n_distinct(artifacts$section05_parcel_shortlist_cluster$zone_id)`
- Parcel-to-zone assignment rate (cluster): `r scales::percent(section_05_coverage$parcel_to_zone_assignment_rate_cluster, accuracy = 0.1)`

Assignment-rate definition: share of retail-flagged parcels that are successfully assigned into a cluster zone within the current market scope.

```{r}
#| label: s05-caveat-note

if (length(section_05_warnings) > 0) {
  tibble::tibble(
    caveat = section_05_warnings
  ) %>%
    gt::gt() %>%
    gt::tab_header(title = "Section 05 Caveats")
}
```

```{r}
#| label: s05-market-context-map
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Market parcel context map showing residential/other and retail parcels before zone filtering. Source: Section 05 + Section 02 context artifacts."

s05_visuals$market_parcel_context_map
```

```{r}
#| label: s05-cluster-context-map
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Cluster zones over parcel context. Source: Section 04 clusters + Section 05 + Section 02 context artifacts."

s05_visuals$cluster_parcel_overlay_map
```

```{r}
#| label: s05-shortlist-cluster-map
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Cluster-system parcel shortlist map. Source: Section 05 visual artifacts."

s05_visuals$shortlist_map_cluster
```

```{r}
#| label: s05-shortlist-cluster-table

datatable_df <- s05_visuals$shortlist_table_cluster %>%
  mutate(
    across(c(parcel_area_sqmi, assessed_value_psf, zone_quality_score, local_retail_context_score, parcel_characteristics_score, shortlist_score), ~ round(.x, 3))
  )

DT::datatable(
  datatable_df,
  filter = "top",
  rownames = FALSE,
  options = list(
    pageLength = 15,
    lengthMenu = c(10, 15, 25, 50, 100),
    scrollX = TRUE,
    autoWidth = TRUE
  ),
  colnames = c(
    "Cluster zone",
    "Rank in zone",
    "Parcel ID",
    "Owner",
    "Owner address",
    "Site address",
    "Use code",
    "Use code description",
    "Use code group",
    "Subtype",
    "Area (sq mi)",
    "Just value",
    "Value / sqft",
    "Recent sale date",
    "Recent sale price",
    "Zone quality",
    "Local context",
    "Parcel score",
    "Final score"
  )
) %>%
  DT::formatCurrency(c("just_value_clean", "last_sale_price"), currency = "$", digits = 0) %>%
  DT::formatRound(
    columns = c("parcel_area_sqmi", "assessed_value_psf", "zone_quality_score", "local_retail_context_score", "parcel_characteristics_score", "shortlist_score"),
    digits = 3
  )
```

# Conclusion and Appendix

This section consolidates top zone highlights, shortlist summary statistics, recommended actions, and QA caveats.
Appendix tables below are the authoritative location for assumptions, caveats, and section-level QA status in this MVP.

```{r}
#| label: s06-zone-highlights

s06_visuals$conclusion_summary_table
```

```{r}
#| label: s06-shortlist-summary

s06_visuals$shortlist_summary_table
```

```{r}
#| label: s06-recommendations

s06_visuals$recommendations_table
```

## Future Deep Dives

Near-term:

- Retail density diagnostics by corridor and by cluster.
- Sensitivity tests on shortlist weights and hybrid quota parameters.
- Parcel-assignment boundary sensitivity (strict in-zone vs buffered assignment).

Long-term:

- Corridor-first clustering seeded from observed retail corridors.
- Alternate parcel marketability signals beyond value-per-sqft.
- Scenario comparisons for different operator formats and target tenant mixes.

## Appendix

Runtime policy:

- This notebook runs in artifact mode only.
- It reads prebuilt `*.rds` outputs and static assets for presentation.
- It does not execute section build, visuals, or checks scripts during render.

Section 3 diagnostics:

```{r}
#| label: s03-score-hist-appendix
#| fig-height: 4
#| fig-width: 9
#| fig-cap: "Distribution of tract scores across all tracts (diagnostic). Source: Section 03 artifacts."

s03_visuals$score_hist_plot
```

```{r}
#| label: s05-zone-quality-appendix
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Cluster zone quality map (mean tract score percentile). Source: Section 05 visual artifacts."

s05_visuals$overlay_map_cluster
```

```{r}
#| label: s06-qa-summary

s06_visuals$qa_summary_table
```

```{r}
#| label: s06-assumptions

s06_visuals$assumptions_caveats_table
```
