---
title: "Retail Opportunity Finder MVP"
subtitle: "Integration Scaffold (Artifact-Driven)"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

## Runtime Policy

This notebook runs in artifact mode only. It reads prebuilt `*.rds` outputs and static assets for presentation.
It does not execute section build, visuals, or checks scripts during render.

```{r}
#| label: f2-load-artifacts
#| include: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(readr)
  library(ggplot2)
  library(gt)
  library(sf)
})

resolve_project_root <- function() {
  candidates <- c(".", "..", "../..", "../../..", "../../../..", "../../../../..")
  for (cand in candidates) {
    root <- normalizePath(cand, winslash = "/", mustWork = FALSE)
    marker <- file.path(root, "notebooks/retail_opportunity_finder/sections")
    if (dir.exists(marker)) return(root)
  }
  stop("Unable to resolve project root from current render working directory.", call. = FALSE)
}

project_root <- resolve_project_root()

resolve_artifact_path <- function(path) {
  if (file.exists(path)) return(path)
  rooted <- file.path(project_root, path)
  if (file.exists(rooted)) return(rooted)
  path
}

read_artifact <- function(path, label) {
  resolved <- resolve_artifact_path(path)
  if (!file.exists(resolved)) {
    stop(paste0("Missing required artifact [", label, "]: ", path), call. = FALSE)
  }
  readRDS(resolved)
}

artifact_manifest <- tribble(
  ~alias, ~path,
  "section01_run_metadata", "notebooks/retail_opportunity_finder/sections/01_setup/outputs/section_01_run_metadata.rds",
  "section01_foundation", "notebooks/retail_opportunity_finder/sections/01_setup/outputs/section_01_foundation.rds",

  "section02_visual_objects", "notebooks/retail_opportunity_finder/sections/02_market_overview/outputs/section_02_visual_objects.rds",
  "section02_validation_report", "notebooks/retail_opportunity_finder/sections/02_market_overview/outputs/section_02_validation_report.rds",

  "section03_funnel_counts", "notebooks/retail_opportunity_finder/sections/03_eligibility_scoring/outputs/section_03_funnel_counts.rds",
  "section03_scored_tracts", "notebooks/retail_opportunity_finder/sections/03_eligibility_scoring/outputs/section_03_scored_tracts.rds",
  "section03_top_tracts", "notebooks/retail_opportunity_finder/sections/03_eligibility_scoring/outputs/section_03_top_tracts.rds",
  "section03_visual_objects", "notebooks/retail_opportunity_finder/sections/03_eligibility_scoring/outputs/section_03_visual_objects.rds",
  "section03_validation_report", "notebooks/retail_opportunity_finder/sections/03_eligibility_scoring/outputs/section_03_validation_report.rds",

  "section04_zone_summary", "notebooks/retail_opportunity_finder/sections/04_zones/outputs/section_04_zone_summary.rds",
  "section04_visual_objects", "notebooks/retail_opportunity_finder/sections/04_zones/outputs/section_04_visual_objects.rds",
  "section04_cluster_zone_summary", "notebooks/retail_opportunity_finder/sections/04_zones/outputs/section_04_cluster_zone_summary.rds",
  "section04_cluster_visual_objects", "notebooks/retail_opportunity_finder/sections/04_zones/outputs/section_04_cluster_visual_objects.rds",
  "section04_cluster_vs_contiguity", "notebooks/retail_opportunity_finder/sections/04_zones/outputs/section_04_cluster_vs_contiguity_comparison.rds",
  "section04_validation_report", "notebooks/retail_opportunity_finder/sections/04_zones/outputs/section_04_validation_report.rds",
  "section04_cluster_validation_report", "notebooks/retail_opportunity_finder/sections/04_zones/outputs/section_04_cluster_validation_report.rds",

  "section05_zone_overlay_contiguity", "notebooks/retail_opportunity_finder/sections/05_parcels/outputs/section_05_zone_overlay_contiguity.rds",
  "section05_zone_overlay_cluster", "notebooks/retail_opportunity_finder/sections/05_parcels/outputs/section_05_zone_overlay_cluster.rds",
  "section05_parcel_shortlist_contiguity", "notebooks/retail_opportunity_finder/sections/05_parcels/outputs/section_05_parcel_shortlist_contiguity.rds",
  "section05_parcel_shortlist_cluster", "notebooks/retail_opportunity_finder/sections/05_parcels/outputs/section_05_parcel_shortlist_cluster.rds",
  "section05_visual_objects", "notebooks/retail_opportunity_finder/sections/05_parcels/outputs/section_05_visual_objects.rds",
  "section05_validation_report", "notebooks/retail_opportunity_finder/sections/05_parcels/outputs/section_05_validation_report.rds",

  "section06_conclusion_payload", "notebooks/retail_opportunity_finder/sections/06_conclusion_appendix/outputs/section_06_conclusion_payload.rds",
  "section06_appendix_payload", "notebooks/retail_opportunity_finder/sections/06_conclusion_appendix/outputs/section_06_appendix_payload.rds",
  "section06_visual_objects", "notebooks/retail_opportunity_finder/sections/06_conclusion_appendix/outputs/section_06_visual_objects.rds",
  "section06_validation_report", "notebooks/retail_opportunity_finder/sections/06_conclusion_appendix/outputs/section_06_validation_report.rds"
)

artifacts <- lapply(seq_len(nrow(artifact_manifest)), function(i) {
  read_artifact(artifact_manifest$path[[i]], artifact_manifest$alias[[i]])
})
names(artifacts) <- artifact_manifest$alias

# Section-level aliases for clean downstream chunks
run_meta <- artifacts$section01_run_metadata
foundation <- artifacts$section01_foundation

s02_visuals <- artifacts$section02_visual_objects
s03_visuals <- artifacts$section03_visual_objects
s04_visuals <- artifacts$section04_visual_objects
s04_cluster_visuals <- artifacts$section04_cluster_visual_objects
s05_visuals <- artifacts$section05_visual_objects
s06_visuals <- artifacts$section06_visual_objects

section_02_report <- artifacts$section02_validation_report
section_03_report <- artifacts$section03_validation_report
section_04_report <- artifacts$section04_validation_report
section_04_cluster_report <- artifacts$section04_cluster_validation_report
section_05_report <- artifacts$section05_validation_report
section_06_report <- artifacts$section06_validation_report

# Defensive validation guards
pass_02 <- all(vapply(section_02_report$checks, function(x) isTRUE(x$pass), logical(1))) &&
  all(unlist(section_02_report$logic_checks))
pass_03 <- all(vapply(section_03_report$checks[1:5], function(x) isTRUE(x$pass), logical(1))) &&
  isTRUE(section_03_report$checks$geometry_check$pass) &&
  all(unlist(section_03_report$logic_checks))

if (!isTRUE(pass_02)) stop("Section 02 validation guard failed.", call. = FALSE)
if (!isTRUE(pass_03)) stop("Section 03 validation guard failed.", call. = FALSE)
if (!isTRUE(section_04_report$pass)) stop("Section 04 validation guard failed.", call. = FALSE)
if (!isTRUE(section_04_cluster_report$pass)) stop("Section 04 cluster validation guard failed.", call. = FALSE)
if (!isTRUE(section_05_report$pass)) stop("Section 05 validation guard failed.", call. = FALSE)
if (!isTRUE(section_06_report$pass)) stop("Section 06 validation guard failed.", call. = FALSE)

section_04_warnings <- if (!is.null(section_04_report$warnings)) section_04_report$warnings else character()
section_05_warnings <- if (!is.null(section_05_report$warnings)) section_05_report$warnings else character()
section_05_coverage <- section_05_report$coverage_metrics

required_visual_keys <- list(
  s02_visuals = c("tiles_layout", "peer_gt", "benchmark_gt", "pop_trend_plot", "distribution_plot"),
  s03_visuals = c("funnel_gt", "price_hist_plot", "growth_hist_plot", "eligible_map_plot", "score_hist_plot", "growth_density_scatter", "top_tracts_gt"),
  s04_visuals = c("zone_map_plot", "zone_summary_gt"),
  s04_cluster_visuals = c("cluster_zone_map_plot", "cluster_summary_gt", "comparison_gt"),
  s05_visuals = c("overlay_map_contiguity", "overlay_map_cluster", "shortlist_map_contiguity", "shortlist_map_cluster", "shortlist_table_contiguity", "shortlist_table_cluster", "system_comparison_gt"),
  s06_visuals = c("conclusion_summary_table", "shortlist_summary_table", "qa_summary_table", "assumptions_caveats_table", "recommendations_table")
)

visual_lists <- list(
  s02_visuals = s02_visuals,
  s03_visuals = s03_visuals,
  s04_visuals = s04_visuals,
  s04_cluster_visuals = s04_cluster_visuals,
  s05_visuals = s05_visuals,
  s06_visuals = s06_visuals
)

for (nm in names(required_visual_keys)) {
  missing_keys <- setdiff(required_visual_keys[[nm]], names(visual_lists[[nm]]))
  if (length(missing_keys) > 0) {
    stop(paste0("Missing visual object keys in ", nm, ": ", paste(missing_keys, collapse = ", ")), call. = FALSE)
  }
}

runtime_manifest <- artifact_manifest %>% mutate(exists = TRUE)

integration_outputs_dir <- file.path(project_root, "notebooks/retail_opportunity_finder/integration/outputs")
if (!dir.exists(integration_outputs_dir)) {
  dir.create(integration_outputs_dir, recursive = TRUE)
}
readr::write_csv(runtime_manifest, file.path(integration_outputs_dir, "phase_f2_runtime_manifest.csv"))
```

## Scaffold Status

```{r}
#| label: f2-status

status_tbl <- tibble::tibble(
  checkpoint = c(
    "Artifact manifest loaded",
    "Section validation guards",
    "Visual object key guards",
    "Runtime mode"
  ),
  status = c("PASS", "PASS", "PASS", "ARTIFACT-ONLY"),
  detail = c(
    paste0(nrow(runtime_manifest), " artifacts loaded via readRDS"),
    "Sections 02-06 validation checks passed",
    "All required visual object names present",
    "No section build/visual/check scripts executed in notebook"
  )
)

status_tbl %>% gt::gt()
```

## Section 1 Setup and Method

This MVP notebook identifies retail corridor opportunity in Jacksonville using a funnel workflow:
funnel screening, tract ranking, zone definition, and parcel shortlist generation.

```{r}
#| label: s01-run-metadata

run_meta_tbl <- tibble::tibble(
  field = c("Run timestamp", "R version", "Git hash", "Target CBSA", "Target year"),
  value = c(
    run_meta$run_timestamp,
    run_meta$r_version,
    run_meta$git_hash,
    run_meta$target_cbsa,
    as.character(run_meta$target_year)
  )
)

run_meta_tbl %>%
  gt::gt() %>%
  gt::tab_header(title = "Run Metadata")
```

```{r}
#| label: s01-model-params

model_params_tbl <- tibble::tibble(
  parameter = c(
    "Minimum growth gate percentile",
    "Maximum price proxy percentile",
    "Maximum density percentile",
    "Top N tracts for ranking output",
    "Score weights (growth/units/headroom/price/commute)"
  ),
  value = c(
    as.character(foundation$model_params$min_growth_percentile),
    as.character(foundation$model_params$price_proxy_pctl_max),
    as.character(foundation$model_params$max_density_percentile),
    as.character(foundation$model_params$top_n_tracts),
    paste(as.character(foundation$model_params$weights), collapse = " / ")
  )
)

model_params_tbl %>%
  gt::gt() %>%
  gt::tab_header(title = "Model Configuration Snapshot")
```

```{r}
#| label: s01-kpi-dictionary

foundation$kpi_dictionary %>%
  gt::gt() %>%
  gt::tab_header(title = "KPI Dictionary (V1)")
```

## Section 2 Market Overview

Jacksonville’s metro profile is presented as a pre-screen before tract-level filtering.
This section keeps the view concise: snapshot KPIs, peer context, benchmark context, and broad U.S. metro placement.
Unless otherwise noted, population growth is shown as share/percent, housing cost metrics are in USD, and supply is units per 1,000 residents.

```{r}
#| label: s02-kpi-tiles

s02_visuals$tiles_layout
```

```{r}
#| label: s02-peer-table

s02_visuals$peer_gt
```

```{r}
#| label: s02-benchmark-table

s02_visuals$benchmark_gt
```

```{r}
#| label: s02-pop-trend
#| fig-height: 5
#| fig-width: 10
#| fig-cap: "Indexed population trend (baseline = 100). Source: Section 02 artifacts derived from `cbsa_features`."

s02_visuals$pop_trend_plot
```

```{r}
#| label: s02-distribution
#| fig-height: 6
#| fig-width: 11
#| fig-cap: "Jacksonville position vs all metro distributions by key metrics. Source: Section 02 artifacts."

s02_visuals$distribution_plot
```

## Section 3 Eligibility and Scoring

The tract screening model narrows the universe through gate logic, then ranks eligible tracts with weighted standardized components.
Current run funnel: `r artifacts$section03_funnel_counts$tracts_remaining[1]` total tracts to
`r artifacts$section03_funnel_counts$tracts_remaining[5]` eligible tracts.
Scoring weights used: growth `0.40`, units `0.25`, headroom `0.20`, price `0.10`, commute `0.05`.

```{r}
#| label: s03-funnel

s03_visuals$funnel_gt
```

```{r}
#| label: s03-price-hist
#| fig-height: 4
#| fig-width: 9
#| fig-cap: "Price proxy percentile distribution with eligibility threshold context. Source: Section 03 artifacts."

s03_visuals$price_hist_plot
```

```{r}
#| label: s03-growth-hist
#| fig-height: 4
#| fig-width: 9
#| fig-cap: "5-year tract population growth distribution. Source: Section 03 artifacts."

s03_visuals$growth_hist_plot
```

```{r}
#| label: s03-eligible-map
#| fig-height: 6
#| fig-width: 9
#| fig-cap: "Eligible tract map after gate logic. Source: Section 03 artifacts."

s03_visuals$eligible_map_plot
```

```{r}
#| label: s03-score-hist
#| fig-height: 4
#| fig-width: 9
#| fig-cap: "Distribution of tract scores among eligible tracts. Source: Section 03 artifacts."

s03_visuals$score_hist_plot
```

```{r}
#| label: s03-growth-density
#| fig-height: 5
#| fig-width: 9
#| fig-cap: "Growth vs density diagnostic with high-scoring tracts highlighted. Source: Section 03 artifacts."

s03_visuals$growth_density_scatter
```

```{r}
#| label: s03-top-tracts

s03_visuals$top_tracts_gt
```

## Section 4 Zone Systems

Cluster zones are the primary narrative system for this MVP because they produce a tighter, decision-usable zone count.
Contiguity outputs are retained for comparison and QA transparency.

Cluster summary for current run:
- Contiguity zones: `r nrow(artifacts$section04_zone_summary)`
- Cluster zones: `r nrow(artifacts$section04_cluster_zone_summary)`

```{r}
#| label: s04-caveat-note

if (length(section_04_warnings) > 0) {
  caveat_tbl <- tibble::tibble(
    caveat = section_04_warnings
  )
  caveat_tbl %>%
    gt::gt() %>%
    gt::tab_header(title = "Section 04 Caveat")
}
```

```{r}
#| label: s04-cluster-map
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Primary (cluster) zone map for MVP narrative. Source: Section 04 cluster artifacts."

s04_cluster_visuals$cluster_zone_map_plot
```

```{r}
#| label: s04-cluster-summary

s04_cluster_visuals$cluster_summary_gt
```

```{r}
#| label: s04-system-comparison

s04_cluster_visuals$comparison_gt
```

### Contiguity Comparison (Secondary)

```{r}
#| label: s04-contiguity-map
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Secondary contiguity zone map for method comparison. Source: Section 04 artifacts."

s04_visuals$zone_map_plot
```

```{r}
#| label: s04-contiguity-summary

s04_visuals$zone_summary_gt
```

## Section 5 Retail Overlay and Shortlist

Section 05 connects zone-level opportunity to parcel-level screening outputs.
The shortlist score combines zone quality, local retail context, and parcel characteristics.

Current run snapshot:
- Cluster shortlist candidates: `r nrow(artifacts$section05_parcel_shortlist_cluster)`
- Covered cluster zones: `r dplyr::n_distinct(artifacts$section05_parcel_shortlist_cluster$zone_id)`
- Parcel-to-zone assignment rate (cluster): `r scales::percent(section_05_coverage$parcel_to_zone_assignment_rate_cluster, accuracy = 0.1)`
- Parcel-to-zone assignment rate (contiguity): `r scales::percent(section_05_coverage$parcel_to_zone_assignment_rate_contiguity, accuracy = 0.1)`

```{r}
#| label: s05-caveat-note

if (length(section_05_warnings) > 0) {
  tibble::tibble(
    caveat = section_05_warnings
  ) %>%
    gt::gt() %>%
    gt::tab_header(title = "Section 05 Caveats")
}
```

```{r}
#| label: s05-overlay-cluster-map
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Cluster-zone retail overlay context map. Source: Section 05 visual artifacts."

s05_visuals$overlay_map_cluster
```

```{r}
#| label: s05-shortlist-cluster-map
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Cluster-system parcel shortlist map. Source: Section 05 visual artifacts."

s05_visuals$shortlist_map_cluster
```

```{r}
#| label: s05-shortlist-cluster-table

s05_visuals$shortlist_table_cluster
```

### Contiguity Comparison (Secondary)

```{r}
#| label: s05-overlay-contiguity-map
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Contiguity-zone retail overlay context map. Source: Section 05 visual artifacts."

s05_visuals$overlay_map_contiguity
```

```{r}
#| label: s05-shortlist-contiguity-map
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Contiguity-system parcel shortlist map. Source: Section 05 visual artifacts."

s05_visuals$shortlist_map_contiguity
```

```{r}
#| label: s05-shortlist-contiguity-table

s05_visuals$shortlist_table_contiguity
```

```{r}
#| label: s05-system-comparison

s05_visuals$system_comparison_gt
```

## Section 6 Conclusion and Appendix

This section consolidates top zone highlights, shortlist summary statistics, recommended actions, and QA caveats.
Appendix tables below are the authoritative location for assumptions, caveats, and section-level QA status in this MVP.

```{r}
#| label: s06-zone-highlights

s06_visuals$conclusion_summary_table
```

```{r}
#| label: s06-shortlist-summary

s06_visuals$shortlist_summary_table
```

```{r}
#| label: s06-recommendations

s06_visuals$recommendations_table
```

### Appendix

```{r}
#| label: s06-qa-summary

s06_visuals$qa_summary_table
```

```{r}
#| label: s06-assumptions

s06_visuals$assumptions_caveats_table
```
